---
title: "10-Regression Analysis with Time Series"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../style.css
params:
  date: !r Sys.Date()    
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following notes and scripts are based on the following sources: [Introductory Econometrics - A Modern Approach](https://www.amazon.com/Introductory-Econometrics-Modern-Approach-Standalone/dp/130527010X/ref=sr_1_2?dchild=1&keywords=Introductory+Econometrics%3A+A+Modern+Approach&qid=1597005903&s=books&sr=1-2) by Jeffrey M. Wooldridge is the main text for the statistical content. This text will be referenced as `(Wooldridge)`.

The companion text [Using R for Introductory Econometrics](http://www.urfie.net/index.html) by Florian Heiss provides specific R scripts in support of the main text and the inspiration for my scripts, notes and graphics. I will reference this document as `(Heiss)`. The following is from **Part 2 - Regression Analysis with Time Series Data, Chapter 10 Basic Regression Analysis with Time Series Data** of `(Heiss)`.      
</div>  

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(data.table)
```

# 1 Introduction

## 1.1 
