---
title: "10-Regression Analysis with Time Series"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../../style.css
params:
  date: !r Sys.Date()    
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following notes and scripts are based on the following sources: [Introductory Econometrics - A Modern Approach](https://www.amazon.com/Introductory-Econometrics-Modern-Approach-Standalone/dp/130527010X/ref=sr_1_2?dchild=1&keywords=Introductory+Econometrics%3A+A+Modern+Approach&qid=1597005903&s=books&sr=1-2) by Jeffrey M. Wooldridge is the main text for the statistical content. This text will be referenced as `(Wooldridge)`.

The companion text [Using R for Introductory Econometrics](http://www.urfie.net/index.html) by Florian Heiss provides specific R scripts in support of the main text and the inspiration for my scripts, notes and graphics. I will reference this document as `(Heiss)`. The following is from **Part 2 - Regression Analysis with Time Series Data, Chapter 10 Basic Regression Analysis with Time Series Data** of `(Heiss)`.      
</div>  

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(data.table)
library(xts)
library(zoo)
library(dynlm)
library(wooldridge)
library(RregressPkg)
library(RplotterPkg)
library(RtsaPkg)
```

# 10-Regression Analysis with Time Series

## 10.1 Static Time Series Models
### Effects of Inflation and Deficits on Interest Rates
1. Set up the data:
```{r}
data("intdef", package = "wooldridge")
intdef_dt <- data.table::setDT(intdef)
intdef_dt <- intdef_dt[,.(i3, inf, def)]
```
2. Define the model formula:
```{r}
formula_obj <- i3 ~ inf + def
```
3. Compute the fitted values and show the coefficient values:
```{r}
intdef_ols <- RregressPkg::ols_calc(
  df = intdef_dt,
  formula_obj = formula_obj
)
RplotterPkg::create_table(x = intdef_ols$coef_df, caption = "Inflation and Deficits on Interest Rates")
```
4. Plot the reponse and predictor variables:
```{r, fig.width = 13, fig.height = 8}
RplotterPkg::multi_scatter_plot(
  df = intdef_dt,
  variables = c("inf", "def"),
  aes_y = "i3",
  col_width = 6,
  row_height = 6,
  title = "Inflation and Deficits on Interest Rates"
)
```
## 10.2 Time Series Data Types in R
### 10.2.1 Equispaced Time Series in R
1. Set up the data:
```{r}
data(barium, package = "wooldridge")
barium_dt <- data.table::setDT(barium)
barium_dt <- barium_dt[,.(chnimp)]
```
2. Create a time series (*ts*) object:
```{r}
imports_ts <- ts(data = barium_dt$chnimp, start = c(1978,2), frequency=12)
RtsaPkg::get_series_info(imports_ts)
```
3. Plot the time series:
```{r}
imports_ts_df <- RtsaPkg::ts_to_df(imports_ts, col_name = "imports")
RplotterPkg::create_scatter_plot(
  df = imports_ts_df,
  aes_x = "DateTime",
  aes_y = "imports",
  x_major_date_breaks = "1 year",
  x_date_labels = "%Y",
  connect = T,
  show_pts = F,
  x_title = "Time"
)
```

### 10.2.2 Irregular Time Series in R
1. Create time series *zoo* object of the interest rate data:
```{r}
data("intdef", package = "wooldridge")
int_year_dt <- data.table::setDT(intdef)
int_year_dt <- int_year_dt[,.(i3, year)]

int_year_zoo <- zoo::zoo(int_year_dt, order.by = int_year_dt$year, frequency = 12)
RtsaPkg::get_series_info(series = int_year_zoo)
```
2. Plot the interest across time:
```{r, fig.width=12, fig.height=8}
int_year_zoo_df <- RtsaPkg::xts_to_df(int_year_zoo, col_names = c("Date", "Interest","Year"))
RplotterPkg::create_scatter_plot(
  df = int_year_zoo_df,
  aes_x = "Date",
  aes_y = "Interest",
  x_major_breaks = seq.Date(from = as.Date("1950", "%Y"), to = as.Date("2000", "%Y"), by = "10 year"),
  x_date_labels = "%Y",
  y_major_breaks = seq(2,14,2),
  connect = T,
  show_pts = F,
  title = "Interest Rate (3-month T-bills)"
)
```

<div class="task">Task:  Plot closing prices on Apple stock from 2000 to latest year.</div>
1. Use the *quantmod::* package to get the data from Yahoo as the source:
```{r}
apple_stocks_xts <- RtsaPkg::get_yahoo_stock_prices(
  symbols = "AAPL",
  from = "2000-01-01",
  to = "2021-01-01",
  periodicity = "daily",
  return_class = "xts"
)
RtsaPkg::get_series_info(series = apple_stocks_xts)
```

2. Plot the data:
```{r, fig.width=13, fig.height=8}
apple_stocks_df <- RtsaPkg::xts_to_df(apple_stocks_xts)
RplotterPkg::create_scatter_plot(
  df = apple_stocks_df,
  aes_x = "index",
  aes_y = "AAPL.Adjusted",
  x_major_date_breaks = "1 year",
  x_date_labels = "%Y",
  x_limits = c(as.Date("2000", "%Y"), as.Date("2021", "%Y")),
  y_major_breaks = seq(0,140,20),
  y_limits = c(0,140),
  x_title = NULL,
  y_title = NULL,
  title = "Apple Adjusted Closing Stock Prices",
  subtitle = "Source: Yahoo",
  connect = T,
  show_pts = F
)
```

## 10.3 Other Time Series Models
### 10.3.1 The *dynlm::* Package
### 10.3.2 Finite Distributed Lag Models

<div class="task">Task: Model Effects of Personal Tax Exemption on Fertility Rates</div>
1. Set up the data:
```{r}
data("fertil3", package = "wooldridge")
```

2. Define yearly time series beginning in 1913:
```{r}
fertil3_ts <- ts(fertil3, start = 1913)
RtsaPkg::get_series_info(fertil3_ts)
```
3. Define linear regression model with lags using *dynlm::dynlm()*:
```{r}
fertil_3_lm <- dynlm::dynlm(gfr ~ pe + L(pe) + L(pe,2) + ww2 + pill, data = fertil3_ts)
```

4. Summarize *fertil_3_lm*:
```{r}
summary(fertil_3_lm)
```
5. Perform an $F$-test with the null hypothesis that the lag coefficients equal zero:
```{r}
car::linearHypothesis(fertil_3_lm,c("pe", "L(pe)", "L(pe, 2)"))
```
<div class="task">Task: Repeat the OLS estimate by lagging the data and using *Rregress::ols_calc()*</div>
1. Convert the data to a *data.table*:

```{r}
data("fertil3", package = "wooldridge")
fertil3_dt <- data.table::setDT(fertil3)
fertil3_dt <- fertil3_dt[, .(gfr, pe, ww2, pill)]
```

2. Lag the *pe* predictor by a factor of 2:
```{r}
fertil3_dt[, c("pe_lag1", "pe_lag2") := shift(pe, 1:2, type = "lag")]
fertil3_dt <- fertil3_dt[!is.na(pe_lag2)]
```

3. Compute the OLS:
```{r}
formula_obj <- gfr ~ pe + pe_lag1 + pe_lag2 + ww2 + pill
fertil3_ols <- RregressPkg::ols_calc(df = fertil3_dt, formula_obj = formula_obj)
```

4. Show the OLS coefficients:
```{r}
RplotterPkg::create_table(fertil3_ols$coef_df)
```

### 10.3.3 Trends
<div class="task">Task: Include a time predictor in estimating housing investment from prices.</div>
1. Set up the data:
```{r}
data("hseinv", package = "wooldridge")
hseinv_dt <- data.table::setDT(hseinv)
hseinv_dt <- hseinv_dt[, .(invpc, year, price)]
```

2. Compute the OLS:
```{r}
formula_obj <- log(invpc) ~ year + log(price)
hseinv_ols <- RregressPkg::ols_calc(df = hseinv_dt, formula_obj = formula_obj)
RplotterPkg::create_table(hseinv_ols$coef_df)
```

