---
title: "3-Linear Regression"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../../style.css
params:
  date: !r Sys.Date()    
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
Chapter 3 Linear Regression: the basics from [Data Analysis Using Regression and Multilevel-Hierarchical Models](http://www.stat.columbia.edu/~gelman/arm/)  
</div>  

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(data.table)
library(ggplot2)
library(foreign)
library(here)
library(RregressPkg)
library(RplotterPkg)

current_dir <- here()
```

## 3.1 One predictor
1. Read the data:
```{r}
file_path <- file.path(current_dir, "Data Analysis Using Regression", "kidiq.dta")
kid_score_df <- foreign::read.dta(file_path)
kid_score_dt <- data.table::setDT(kid_score_df)
kid_score_dt <- kid_score_dt[, .(kid_score, mom_hs, mom_iq)]
kid_score_dt[, mom_hs := as.factor(mom_hs)]
```

2. Regress *kid_score* on *mom_hs*:
```{r}
formula_obj <- kid_score ~ mom_hs
one_predict_ols <- RregressPkg::ols_calc(
  df = kid_score_dt,
  formula_obj = formula_obj
)
RplotterPkg::create_table(
  x = one_predict_ols$coef_df
)
```

3. Plot the regression:
```{r}
plot_dt <- data.table(
  hs = kid_score_dt$mom_hs,
  score = kid_score_dt$kid_score,
  fit = one_predict_ols$fitted_vals
)
RplotterPkg::create_scatter_plot(
  df = plot_dt,
  aes_x = "hs",
  aes_y = "score",
  x_title = "Mother completed high school",
  y_title = "Child test score",
  x_major_breaks = c(0,1),
  position = position_jitter(width = 0.01, height = 0.01)
) + geom_line(aes(y = fit), color="blue")
```

## 3.2 Multiple predictors

1.  Regress *kid_score* on *mom_hs* and *mom_iq*:
```{r}
formula_obj <- kid_score ~ mom_hs + mom_iq
multi_predict_ols <- RregressPkg::ols_calc(
  df = kid_score_dt,
  formula_obj = formula_obj
)
RplotterPkg::create_table(
  x = multi_predict_ols$coef_df
)
```

2. Plot the regression grouped by *mom_hs*:
```{r}
plot_dt <- data.table(
  hs = kid_score_dt$mom_hs,
  iq = kid_score_dt$mom_iq,
  score = kid_score_dt$kid_score,
  fit = multi_predict_ols$fitted_vals 
)
RplotterPkg::create_scatter_plot(
  df = plot_dt,
  aes_x = "iq",
  aes_y = "score",
  aes_fill = "hs",
  x_title = "Mother IQ score",
  y_title = "Child test score",
  palette_colors = c("blue","green"),
  pts_size = 2.0
) + geom_line(data = plot_dt[hs == 0], aes(y = fit), color = "blue", alpha = 0.5, size = 2.0) +
  geom_line(data = plot_dt[hs == 1], aes(y = fit), color = "green", alpha = 0.5, size = 2.0)
```

## 3.3 Interactions
1.  Regress *kid_score* on *mom_hs*, *mom_iq* and *mom_hs x mom_iq*:
```{r}
formula_obj <- kid_score ~ mom_hs + mom_iq + mom_hs * mom_iq
inter_predict_ols <- RregressPkg::ols_calc(
  df = kid_score_dt,
  formula_obj = formula_obj
)
RplotterPkg::create_table(
  x = inter_predict_ols$coef_df
)
```

2. Plot the regression grouped by *mom_hs*:
```{r}
plot_dt <- data.table(
  hs = kid_score_dt$mom_hs,
  iq = kid_score_dt$mom_iq,
  score = kid_score_dt$kid_score,
  fit = inter_predict_ols$fitted_vals 
)
RplotterPkg::create_scatter_plot(
  df = plot_dt,
  aes_x = "iq",
  aes_y = "score",
  aes_fill = "hs",
  x_title = "Mother IQ score",
  y_title = "Child test score",
  palette_colors = c("blue","green"),
  pts_size = 2.0
) + geom_line(data = plot_dt[hs == 0], aes(y = fit), color = "blue", alpha = 0.5, size = 2.0) +
  geom_line(data = plot_dt[hs == 1], aes(y = fit), color = "green", alpha = 0.5, size = 2.0)
```

## 3.6 Assumptions and diagnostics

<div class="task">Task: Plot residuals versus fitted values.</div>
```{r}
RregressPkg::plot_fit_residuals(
  df =  kid_score_dt,
  formula_obj =  kid_score ~ mom_iq,
  label_sd = 1.0,
  x_title = "Fitted Mother IQ score",
  y_title = "Model Residuals",
  y_limits = c(-60.0,50.0),
  y_major_breaks = seq(-60, 50, 10)
)
```

 


