---
title: "Testing the Equality of Regression Coefficients"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../style.css
params:
  date: !r Sys.Date()    
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>

Author: Rick Dean

</div>

<div>

Article date: `r params$date`

</div>

::: {.abstract}
<p class="abstract">

Abstract

</p>

The following notes and R scripts were inspired by an article titled [Testing the Equality of Regression Coefficients](https://shouldbewriting.netlify.app/posts/2021-02-16-testing-coefficients-equality/) by Mattan S. Ben-Shachar. The author compares four methods with an example data set.
:::

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(data.table)
library(effectsize)
library(RregressPkg)
library(RplotterPkg)
library(car)
library(ggplot2)
```

## The data set

### Access the data

Source of the data is `effectsize::hardlyworking` where the response is a *worker's salary* (**salary**) and predictors *number of extra hours* (**xtra_hours**) and *number of compliments* (**n_comps**).

```{r}
data("hardlyworking", package = "effectsize")
head(hardlyworking)
```

### Plot the response with the predictors

```{r}
RplotterPkg::multi_scatter_plot(
  df = hardlyworking,
  variables = c("xtra_hours", "n_comps"),
  aes_y = "salary",
  title = "Salary vs Extra Hours and No. of Complements",
  subtitle = "Data set: effectsize::hardlyworking",
  rot_y_tic_label = T,
  col_width = 5
)
```

### Standardize the data

```{r}
hardlyworking_dt <- data.table::setDT(hardlyworking)
hardlyworking_dt[, `:=`(salary_z = scale(salary), xtra_hours_z = scale(xtra_hours), n_comps_z = scale(n_comps))]
```

## Method 1: As model comparisons

### Compute the OLS estimates for the first lm model `salary_z ~ xtra_hours_z + n_comps_z`

```{r}
model_1_lst <- RregressPkg::ols_calc(df = hardlyworking_dt, formula_obj = salary_z ~ xtra_hours_z + n_comps_z)
RplotterPkg::create_table(x = model_1_lst$coef_df, caption = "Model 1: salary_z ~ xtra_hours_z + n_comps_z")
```

### Compute the OLS estimates for the second lm model `salary_z ~ I(xtra_hours_z + n_comps_z)`

```{r}
model_2_lst <- RregressPkg::ols_calc(df = hardlyworking_dt, formula_obj = salary_z ~ I(xtra_hours_z + n_comps_z))
RplotterPkg::create_table(x = model_2_lst$coef_df, caption = "Model 2: salary_z ~ I(xtra_hours_z + n_comps_z)")
```

### Compare the models via ANOVA

Model 1 ANOVA

```{r}
RplotterPkg::create_table(x = model_1_lst$anova_df, caption = "Model 1: salary_z ~ xtra_hours_z + n_comps_z")
```

Model 2 ANOVA

```{r}
RplotterPkg::create_table(x = model_2_lst$anova_df, caption = "Model 2: salary_z ~ I(xtra_hours_z + n_comps_z)")
```


Using `stats::anova()`:
```{r}
 model_1_lm <- lm(salary_z ~ xtra_hours_z + n_comps_z, data = hardlyworking_dt)
 model_2_lm <- lm(salary_z ~ I(xtra_hours_z + n_comps_z), data=hardlyworking_dt)
 stats::anova(model_1_lm,model_2_lm)
```

::: {.note}
Note: It appears that model 1 is preferred (with a lower SS Residual Error) and its predictors are showing that $\hat{\beta}_{XtraHours}$ \> $\hat{\beta}_{Compliments}$
:::

## Method 2: Paternoster et al (1998)

Compute the $t$-test to compare the coefficients according to [Paternoster et al (1998)](https://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.1025.8565&rep=rep1&type=pdf). From the first model results that separates the two coefficients (*model_1\_lst*): 1. The model's coefficient differences:

```{r}
b_x_hours <- model_1_lst$coef_df$Value[[2]]
b_n_comp <- model_1_lst$coef_df$Value[[3]]
diff_coef =b_x_hours - b_n_comp
```
diff_coef = `r diff_coef`

2.  The model's standard errors of the coefficients:

```{r}
sd_x_hours <- model_1_lst$coef_df$SE[[2]]
sd_n_comp <- model_1_lst$coef_df$SE[[3]]
```
sd_x\_hours = `r sd_x_hours`

sd_n\_comp = `r sd_n_comp`

3.  Use the correct formula for computing the $t$-statistic for the difference in coefficients: $$Z = \frac{b_{1} - b_{2}}{\sqrt{SEb_{1}^2 + SEb_{2}^2}}$$
```{r}
Z = diff_coef / sqrt(sd_x_hours^2 + sd_n_comp^2)
```
Z = `r Z`

4. The variance for the difference between two coefficients:
$$Var(\hat\beta_{1} - \hat\beta_{2}) = Var(\hat\beta_{1}) + Var(\hat\beta_{2}) - 2*Cov(\hat\beta_{1},\hat{\beta_{2}})$$
```{r}
var_diff <- sd_n_comp^2 + sd_x_hours^2 - 2 * model_1_lst$var_cov[2,3]

se_diff <- sqrt(var_diff)
```
var_diff = `r var_diff`

se_diff = `r se_diff`

5. The $t$-statistic with n-p degrees of freedom (n = 500, p = 2 var + 1 intercept) = 497:
$$t_{diff} = \frac{b_{1} - b_{2}}{SE_{diff}}$$
```{r}
t_diff = diff_coef/se_diff
```
t_diff = `r t_diff` 

Z = `r Z`

## Method 5: Using `car::linearHypothesis()`
```{r}
car::linearHypothesis(model_1_lm, c("xtra_hours_z - n_comps_z"))
```






