---
title: "Create a Covariance/Correlation Matrix"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../style.css
params:
  date: !r Sys.Date()    
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following R scripts in manually creating a covariance was inspired by the article [Making a Covariance Matrix in R](https://stats.seandolinar.com/making-a-covariance-matrix-in-r/) by seandolinar(GitHub)
</div>  

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(data.table)
library(RplotterPkg)
```

## The data

```{r}
data_dt <- data.table(
  math = c(84, 82, 81, 89, 73, 94, 92, 70, 88, 95),
  science = c(85, 82, 72, 77, 75, 89, 95, 84, 77, 94),
  history = c(97, 94, 93, 95, 88, 82, 78, 84, 69, 78))

k <- ncol(data_dt)
n <- nrow(data_dt)
```

## Creating the covariance/correlation manually

1. Create a matrix of means for each column:
```{r}
means_dt <- data_dt[,.(math = mean(math), science = mean(science), history = mean(history))]
means_mt <- matrix(data = 1, nrow = n) %*% as.matrix(means_dt, rownames = F)
means_mt
```

2. Subtract *means_mt* from *data_dt*:
```{r}
data_mt <- as.matrix(data_dt, rownames = F)
diff_mt <- data_mt - means_mt
diff_mt
```

3. Create the covariance matrix:
```{r}
var_cov_mt <- t(diff_mt) %*% diff_mt * (n - 1)^-1
var_cov_mt
```

4. Pull variance and standard deviations:
```{r}
var_mt <- diag(var_cov_mt)
sd_mt <- sqrt(var_mt)
```

Variance = `r var_mt`

SD = `r sd_mt`

5. Create the correlation matrix:
```{r}
se_product_mt <- sd_mt %*% t(sd_mt)
corr_mt <- var_cov_mt / se_product_mt
```

## Creating the covariance/correlation via stats:: package

1. Covariance using `stats::cov()`
```{r}
var_cov_R_mt <- stats::cov(data_dt)
```

2. Correlation usine `stats::cor()`
```{r}
corr_R_mt <- stats::cor(data_dt)
```

## Compare correlation results

1. Correlations manually:
```{r}
RplotterPkg::create_table(
  x = corr_mt,
  caption = "Correlations Manually"
)
```

2. Correlations via `stats::cor()`:
```{r}
RplotterPkg::create_table(
  x = corr_R_mt,
  caption = "Correlations via stats::cor()"
)
```

