---
title: "How to Make the Most of Regression: Standardization and Post-Estimation Simulation"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../style.css
params:
  date: !r Sys.Date()    
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following are notes and R scripts inspired by the article [How to Make the Most of Regression: Standardization and Post-Estimation Simulation](http://svmiller.com/blog/2020/04/post-estimation-simulation-trump-vote-midwest/) by Steve Miller.    
</div>  

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(data.table)
library(ggrepel)
library(RplotterPkg)
library(RregressPkg)
library(here)

current_dir <- here()
```

## The data and first model

We'll be using the 2016 Cooperative Congressional Election Study (CCES) data.

1. Load the data:
```{r}
data_path <- file.path(current_dir, "How to Make the Most of Regression/data/TV16.rda")
load(file = data_path)
tv16_dt <-  data.table::setDT(TV16)
```

2. Using a subset of midwest states (Indiana, Michigan, Ohio, Pennsylvania, Wisconsin) amd white respondents:
```{r}
tv16_dt <- tv16_dt[state %in% c("Indiana","Michigan","Ohio","Pennsylvania","Wisconsin") & racef == "White", .(TrumpVote = votetrump, Age = age, Female = female, CollegeEdu = collegeed, HouseholdInc = famincr, Partisanship = pid7na, Ideology = ideo, BornAgainChris = bornagain, CognitiveRace = lcograc, EmpatheticRace = lemprac)]
```

3. Create the first model:
```{r}
model_1 <- stats::glm(TrumpVote ~ Age + Female + CollegeEdu + HouseholdInc + Partisanship + Ideology + BornAgainChris + CognitiveRace + EmpatheticRace, data = tv16_dt, family = binomial(link = "logit"), na.action = na.exclude)
```







