---
title: "Using dbinom and pbinom from R with Baseball Stats"
author: "Rick Dean"
date: "07/18/2020"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../style.css
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 8)
```

The article by Joshua Cook titled Riddler: [Can You Beat MLB Recods?](https://joshuacook.netlify.app/post/riddler-beat-mlb-records/) inspired the content in this R markdown document.

```{r}
library(knitr)
library(ggplot2)
library(data.table)
library(RplotterPkg)
set.seed(123)
```

# 1 First question
<div class="task">If you have four at-bats per game, what are your chances of batting at least .400 over the course of the *60-game* season?</div> 

## 1.1 Solution
The solution is just the probability of observing a batting average of 0.400 or greater. The first value is computed using dbinom() and the second cumulative probability is calculated using pbinom(), setting lower.tail = FALSE to get the tail above 0.400.

1. Observing a batting average of 0.400:
```{r}
num_at_bats <- 60 * 4
real_batting_average <- 0.350
target_batting_average <- 0.400

prob_at_400 <- stats::dbinom(
  x = target_batting_average * num_at_bats,
  size = num_at_bats,
  prob = real_batting_average
  )
```
<div class="note">Note: `stats::dbinom()` gives the density.</div>
The probability of observing a batting average of 0.400: `r prob_at_400`

2. Observing a batting average greater than 0.400:
```{r}
prob_above_400 <- stats::pbinom(
  q = target_batting_average * num_at_bats,
  size = num_at_bats,
  prob = real_batting_average,
  lower.tail = FALSE
)
prob_at_above_400 <- prob_at_400 + prob_above_400
```
<div class="note">Note: `stats::pbinom()` gives the distribution function.</div>

The probability of observing a batting average above 0.400: `r prob_above_400`

Total probability of observing a batting average at or above 0.400: `r prob_at_above_400`

There is a 6.1% chance of reaching a batting average of 0.400 in the shorter season

# 2 Second question
<div class="task">If you have four at-bats per game, what are your chances of batting at least .400 over the course of the *162-game* season?</div> 

## 2.1 Solution
Only the probability of a batting average greater than 0.400 needs to be
calculated (because .400 * 162 * 4 is a non-integer value).

```{r}
num_at_bats <- 162 * 4
real_batting_average <- 0.350
target_batting_average <- 0.400

prob_above_400_162_games <- stats::pbinom(
  q = target_batting_average * num_at_bats,
  size = num_at_bats,
  prob = real_batting_average,
  lower.tail = FALSE
)
```
With 162 games, there is a probability `r prob_above_400_162_games` of achieving a batting average of at least 0.400.

# 3 Simulation solution
The solution to this riddle could also be found by simulating a whole
bunch of seasons with the real batting average of 0.350 and then just
counting how frequently the simulations resulted in an observed batting
average of 0.400.

## 3.1 Solution for 1 simulation (season)
A single season can be simulated using the `stats::rbinom()` function which generates random deviates. Arguments for the function include:

 + n = number of seasons to simulate
 + size = number of at bats
 + prob = true batting average (real_batting_average/num_at_bats)
 
 The returned value is a sampled number of hits ("successes") over the season from the binomial distribution.
 For n = 1, a single season:
```{r}
num_at_bats <- 60 * 4
real_batting_average <- 0.350

hits_1 <- stats::rbinom(
  n = 1,
  size = num_at_bats,
  prob = real_batting_average
)
batting_average_1 <- hits_1/num_at_bats
```
The simulated hits for 1 season: `r hits_1`

The simulated average batting for 1 season: `r batting_average_1`
 
## 3.2 Solution for 10000 simulations (seasons)
Taking n = 10000 simulations
```{r}
num_at_bats <- 60 * 4
real_batting_average <- 0.350

hits_10000 <- stats::rbinom(
  n = 10000, # 10000 simulations
  size = num_at_bats,
  prob = real_batting_average
)
batting_averages <- hits_10000/num_at_bats
batting_average_10000 <- mean(batting_averages)
```
The simulated average batting across 10000 seasons: `r batting_average_10000` which is close to *real_batting_average*.


### 3.2.1 Plot the density of batting averages across 1000 seasons
```{r, fig.width=12}
target_batting_average <- 0.400
batting_averages_dt <- data.table(
  x = batting_averages
)
RplotterPkg::create_density_plot(
  df = batting_averages_dt,
  aes_x = "x",
  title = "Simulated Batting Averages Across 10000 Seasons",
  subtitle = "Real batting average = 0.350; target average = 0.400",
  x_title = "10000 simulated season batting averages",
  y_title = "probability density",
  density_fill = "black",
  density_alpha = 0.2,
  x_limits = c(0.2,0.5),
  x_major_breaks = seq(0.2, 0.5, 0.05),
  y_major_breaks = seq(0.0, 13.5, 1.5),
  rot_y_tic_label = TRUE,
  plot_obs = TRUE,
  plot_obs_jitter = TRUE
) + 
  geom_vline(xintercept = target_batting_average, color="tomato", lty = 2, size = 1.2)
```

### 3.2.2 Probability of a batting average greater than 0.400 from the simulation
```{r}
prob_at_above_400_simulation <- mean(batting_averages >= 0.40)
```
From the simulation the average batting probabilities at or greater than .400 is `r prob_at_above_400_simulation` which is close to the above actual answer of `r prob_at_above_400`.

# 4 Why the length of the season matters to the distribution
The function `stats::rbinom()` generates a vector of binomial distributed random variables given a vector length *n*, number of trials *size*, and probability of success on each trial *prob*.

## 4.1 Generate a single season of "at-bats" using `stats::rbinom()`
4 "at-bats" per game over 60 game season with a batting average of 0.350
```{r}
sampled_season_hits <- stats::rbinom(n = 60*4, size = 1, prob = 0.350)
season_hits_dt <- data.table(
  hits = sampled_season_hits,
  at_bat = 1:length(sampled_season_hits),
  cum_hits = cumsum(sampled_season_hits),
  batting_avg = cumsum(sampled_season_hits)/1:length(sampled_season_hits)
)
```
## 4.2 Plot the running batting average over a single simulated season
```{r, fig.width=12}
RplotterPkg::create_scatter_plot(
  df = season_hits_dt,
  aes_x = "at_bat",
  aes_y = "cum_hits",
  aes_color = "batting_avg",
  title = "Running batting average over a simulated season",
  x_title = "At Bat Number",
  y_title = "Cumulative Hits",
  y_major_breaks = seq(0, 90, 10),
  connect = TRUE,
  show_pts = FALSE,
  line_size = 1.6
) + ggplot2::scale_color_viridis_c() +
  theme(legend.position = c(0.85, 0.35)) +
  labs(color = "batting average")
```

## 4.3 Plot the running batting average over multiple simulated seasons
```{r}

```

