---
title: "PennState Stat 501 Lesson 7 - MLR Estimation & Prediction"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../style.css
params:
  date: !r Sys.Date()    
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>

Author: Rick Dean

</div>

<div>

Article date: `r params$date`

</div>

::: {.abstract}
<p class="abstract">

Abstract

</p>

The following notes, and R scripts are based on the online course [PennState Lesson 7 MLR Estimation and Prediction](https://online.stat.psu.edu/stat501/lesson/7)
:::

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(here)
library(data.table)
library(RregressPkg)
library(RplotterPkg)

current_dir <- here::here("PennState-Stat 501 Regression Methods")
```

## 7.1 Confidence interval for the mean response
A confidence interval reflects the uncertainty around the **mean** response values, while the prediction interval reflects the uncertainty around a **single** response value.

Which one should we use? The answer to this question depends on the context and the purpose of the analysis.  Generally we are interested in specific individual predictions, so a prediction interval would be more appropriate. Using a confidence interval when you should be using a prediction interval will greatly underestimate the uncertainty in a given value.

The standard error of the fit for **in-scope** $X_{h}$ is given by: $$se(\hat{y_{h}}) = \sqrt{MSE(X_{h}^T(X^TX)^{-1}X_{h})}$$ and the confidence interval is:

$$\hat{y}_{h}  \pm  t_{(\alpha/2,n-p)} \times se(\hat{y_{h}})$$ where

a.  The above formula for $\hat{y_{h}}$ is the **fitted value** when predictors $X_{h}$ are within "**scope of the model**. Section 7.2 presents the  **predicted value** when we have a new response $y_{new}.

b.  $t_{(\alpha/2,n-p)}$ is the $t$-multiplier. Note that the $t$-multiplier has $n-p$ degrees of freedom because the confidence interval uses the mean square error (MSE) whose denominator is $n-p$.

## Example using the IQ Size data

### Manually estimate the fit, se, and 95% CI

For $X_{h}$ we have *Brain* = 90.0 and *Height* = 70.0.

1.  Read in the data:

```{r}
data_path <- file.path(current_dir, "data/iqsize.txt")
iqsize_dt <- data.table::fread(data_path)
iqsize_dt <- iqsize_dt[, !c("Weight")]
```

2.  Estimate the OLS parameters:

```{r}
iqsize_ols_lst <- RregressPkg::ols_calc(
  df = iqsize_dt,
  formula_obj = PIQ ~ Brain + Height
)
```

3.  Compute the fitted value $\hat{y}_{h}$:

```{r}
coef_vals <- iqsize_ols_lst$coef_vals
y_h <- coef_vals[["(Intercept)"]] + coef_vals[["Brain"]] * 90.0 + coef_vals[["Height"]] * 70.0
y_h
```

4.  Compute the se of the fit:

```{r}
X <- as.matrix(iqsize_dt[, !c("PIQ")])
Inter_v <- c(Inter = 1)
X <- cbind(Inter_v,X)
x2_inv <- solve(t(X) %*% X)

X_h <- matrix(c(1.0, 90.0, 70.0), ncol = 3)

X_hh <- X_h %*% x2_inv %*% t(X_h)

se_fit <- sqrt(iqsize_ols_lst$mse * X_hh)[,1]
se_fit
```

5.  Compute the 95% CI's for $\hat{y}_{h}$:

```{r}
n = nrow(X)
p = ncol(X)
t = stats::qt(0.975, n - p)
CI_upper <- y_h + t * se_fit
CI_lower <- y_h - t * se_fit
```

CI_upper = `r CI_upper`

CI_lower = `r CI_lower`

The take away:

> We can be 95% confident that the average performance IQ score of all college students with brain size = 90 and height = 70 is between 98.24 and 113.04 counts per 10,000.

### Estimate the confidence intervals using `stats::predict()`

1. Compute the linear model:
```{r}
iqsize_lm <- lm(data = iqsize_dt, formula = PIQ ~ Brain + Height)
iqsize_lm
```
2. Define one set of predictor values:
```{r}
predictors_df <- data.frame(
  Brain = 90.0,
  Height = 70.0
)
```

3. Compute the fit and the 95% confidence interval:
```{r}
iqsize_confidence <- stats::predict(iqsize_lm, predictors_df, interval = "confidence", se.fit = T)
iqsize_confidence
```
4. Estimate the confidence intervals using `RregressPkg::ols_predict_calc()`:
```{r}
iq_confidence_df <- RregressPkg::ols_predict_calc(
  df = iqsize_dt,
  formula_obj = PIQ ~ Brain + Height,
  predictor_vals_df = data.frame(Brain = 90.0, Height = 70.0)
)

RplotterPkg::create_table(
  x = iq_confidence_df,
  caption = "IQ 95% Confidence Interval"
)
```


## 7.2 Prediction interval for a new response 

### Estimate the prediction interval for the IQ Size data

Using `RregressPkg::ols_predict_calc()`.

1. Set up the data set:
```{r}
data_path <- file.path(current_dir, "data/iqsize.txt")
iqsize_dt <- data.table::fread(data_path)
iqsize_dt <- iqsize_dt[, !c("Weight")]

iqsize_lm <- lm(data = iqsize_dt, formula = PIQ ~ Brain + Height)
```

2. Define the predictor values:
```{r}
predictors_df <- data.frame(
  Brain = 90.0,
  Height = 70.0
)
```

3. Compute the fit and the 95% prediction interval:
```{r}
iqsize_prediction <- stats::predict(iqsize_lm, predictors_df, interval = "prediction", se.fit = F)
iqsize_prediction
```

4. Estimate the confidence intervals using `RregressPkg::ols_predict_calc()`:
```{r}
iqsize_prediction_df <- RregressPkg::ols_predict_calc(
  df = iqsize_dt,
  formula_obj = PIQ ~ Brain + Height,
  predictor_vals_df = predictors_df,
  interval = "prediction"
)

RplotterPkg::create_table(x = iqsize_prediction_df, caption = "IQ Predictions")
```



